---
title: "workingwithdata"
format: html
---

# Introduction

## \*Working with data

#install.packages("descr")

```{r}
library(descr)
library(tidyverse)
```

```{r}
file_path<-file.path("..","data","data.csv")
df<-read_csv(file_path)
head(df)

```

```{r}
file_path<-file.path("..","data","codebook.csv")
codebook<-read_csv(file_path)
head(codebook)
```

## \*Data management

I began by loading the dataset and checking for missing or inconsistent values. Then I used the (freq) function to examine the distribution of key trauma-related variables, such as witnessing a friend have a panic attack, being raped by a family member, and being ignored by a parent.

The frequency tables show how often participants reported each experience. For example, most responses indicate “No,” suggesting that only a small portion of participants experienced these specific traumas. This helps me understand how common or rare each event is in the dataset.

### **Witnessed a friend have a panic attack**

```{r}
freq(as.ordered(df$`witnessed a friend have a panic attack`), plot=FALSE)
```

```         
```

### **Trauma raped by a family member**

```{r}
freq(as.ordered(df$`was raped by a family member`), plot=FALSE)

```

### **Trauma by being ignored by a parent**

```{r}
freq(as.ordered(df$`was ignored by a parent`), plot=FALSE)

```

## 

# **\* Univariate Data Visualizations**

```{r}
library(ggplot2)
library(dplyr)
library(papaja)
```

### **Figure 1 Bar Graph Ignored By Parent**

```{r}
# univariate graph (categorical variable)
ggplot(df, aes(x = `was ignored by a parent`)) +
  geom_bar() +
  ggtitle("Trauma Effects")


```

Bar graph shows that most participants reported **not** being ignored by a parent, while a smaller portion indicated that they had experienced this form of emotional neglect. This suggests that parental neglect was less common in this sample, but still present enough to represent a meaningful trauma experience among some participants.

### Figure 2 Raped by How many Participants reported being raped by a family member?

```{r}
# Bar graph for "was raped by a family member"
ggplot(df, aes(x = `was raped by a family member`)) +
  geom_bar(fill = "lightcoral") +
  ggtitle("Reported Trauma: Raped by a Family Member") +
  xlab("Response") +
  ylab("Number of Participants") +
  theme_minimal()
```

The graph shows that the majority of participants reported not being raped by a family member. Only a small portion indicated “Yes.” It tells us that while this specific trauma was relatively rare among respondents, it represents a severe and sensitive form of trauma that requires careful attention in data interpretation.

```{r}
df %>%
  count(`was raped by a family member`) %>%
  mutate(percent = n / sum(n) * 100)
```

The table above shows us how many people answered each category and the percentage.

### Figure 3 How many participants reported witnessing a friend have a paic attack?

```{r}
# Bar graph for "witnessed a friend have a panic attack"
ggplot(df, aes(x = `witnessed a friend have a panic attack`)) +
  geom_bar(fill = "lightblue") +
  ggtitle("Witnessed a Friend Have a Panic Attack") +
  xlab("Response") +
  ylab("Number of Participants") +
  theme_minimal()
```

The bar graph above tells us that only a small percentage of participants witnessed a friend and experiencing a panic attack. Most participants responded "no", its uncommon.
