---
title: "workingwithdata"
format: html
---

# Introduction

## \*Working with data

#install.packages("descr")

```{r}
library(descr)
library(tidyverse)
```

```{r}
file_path<-file.path("..","data","data.csv")
df<-read_csv(file_path)
head(df)
```

### **Witnessed a friend have a panic attack**

```{r}
freq(as.ordered(df$`witnessed a friend have a panic attack`), plot=FALSE)
```

```         
```

### **Trauma raped by a family member**

```{r}
freq(as.ordered(df$`was raped by a family member`), plot=FALSE)

```

### **Trauma by being ignored by a parent**

```{r}
freq(as.ordered(df$`was ignored by a parent`), plot=FALSE)

```

## \*Data management

I began by loading the dataset and checking for missing or inconsistent values. Then I used the (freq) function to examine the distribution of key trauma-related variables, such as witnessing a friend have a panic attack, being raped by a family member, and being ignored by a parent.

The frequency tables show how often participants reported each experience. For example, most responses indicate “No,” suggesting that only a small portion of participants experienced these specific traumas. This helps me understand how common or rare each event is in the dataset.
